The function--sheaf dictionary for smooth commutative group schemes over finite fields

### 2014-08-15 dictionary.tex
I've added the insight using descent to Section 3.
But I’m not finished with Section 3.

\section{The main result}\label{sec:main}

Having found $\QCiso{G}$ and $\Hom(\qcs{L},\qcs{L}')$ when $G$ is an etale commutative group scheme over $\Fq$, we now wish to find these for all commutative group scheme over $\Fq$. The next section introduces a concept that will be helpful in that regard.

@@

Let $G$ be a smooth commutative group scheme over $\Fq$.
In this section we consider a class of quasicharacter sheaves on $G$ obtained by
replacing the Weil sheaf $(\gqcs{L}, \phi)$ on $\bG$ in the definition of a quasicharacter
sheaf with an $\ell$-adic local system on $G$ itself.

@@

We will identify $\bQC(G)$ with its image under $B_G$
and say that a quasicharacter sheaf $\qcs{L} \in \QC(G)$ \emph{descends to a bounded quasicharacter sheaf} on $G$
if it is isomorphic to some $B_G(\qcs{L}_0, \mu_0)$.
We will see in Corollary~\ref{cor:bounded-and-finite} that bounded quasicharacter sheaves are exactly quasicharacter sheaves with bounded Trace of Frobenius.\todo{Check this, later.}

In general, $\bQC(G)$ is an essentially
proper subcategory of $\QC(G)$. 
However, if $G$ is connected, the categories are equivalent, as we now show.

\begin{proposition}\label{prop:bounded_connected}
If $G$ is a connected commutative algebraic group over $\Fq$ then
$B_G : \bQC(G) \to \QC(G)$ is an equivalence of categories.
\end{proposition}

@@

\subsection{Objects, when $G$ is connected and of finite type}\label{sec:connected}

\begin{proposition}\label{prop:obj_connected}
 If $G$ is a connected, commutative algebraic group over $\Fq$ then
 \[
  \TrFrob{G} : \QCiso{G} \to G(\Fq)^*
 \]
 is an isomorphism of groups.
\end{proposition}

\begin{proof}
At its heart, this is Deligne's function--sheaf dictionary for connected commutative algebraic groups over finite fields, as we now briefly recall from \citelist{\cite{deligne:SGA4.5}*{Sommes trig.}, \cite{laumon:87a}*{1.1.3}}.

We begin by showing that $\TrFrob{G} : \QCiso{G} \to G(\Fq)^*$ is surjective.
Let $\chi : G(\Fq) \to \EEx$ be a character.
Let $\Lang : G \to G$ be the Lang isogeny for $G$.
Set $\qcs{L}_0 = ( \Lang_* \EE)_{\chi^{-1}}$, the $\chi^{-1}$-isotypic component of the push-forward of the constant sheaf on $G$ along the Lang isogeny.
Let $\gqcs{L}$ be the local system on $\bG$ obtained by pull-back along $\bG \to G$.
Then $\gqcs{L}$ comes equipped with an isomorphism $\phi : \Frob{G}^* \gqcs{L} \to \gqcs{L}$.
As explained in \cite{laumon:87a}*{1.1.3}, the trace of the Weil sheaf $(\gqcs{L},\phi)$ is $\chi$.
The sheaf $\qcs{L}_0$ also comes equipped with an isomorphism 
$\mu_0 : m^*\qcs{L}_0 \to \qcs{L}_0\boxtimes \qcs{L}_0$ 
\cite{laumon:87a}*{1.1.3.2},
which satisfies the analogue of \ref{QC.2} over $G$.
Writing $\mu : \bm^* \gqcs{L}  \to \gqcs{L} \boxtimes \gqcs{L}$ for the isomorphism obtained by base change, it follows immediately that $\mu$ satisfies conditions \ref{QC.2} and \ref{QC.3}.
Thus, $\qcs{L}\ceq (\gqcs{L},\mu,\phi)$ is a quasicharacter sheaf on $G$ with $\trFrob{\qcs{L}} = \chi$.
It follows that $\TrFrob{G} : \QCiso{G} \to G(\Fq)^*$ is surjective.

Now, let $\qcs{L}$ and $\qcs{L}'$ be quasicharacter sheaves on $G$ with $\trFrob{\qcs{L}} = \trFrob{\qcs{L}'}$. We will show $\qcs{L} \iso \qcs{L}$.
We have just seen that $\qcs{L}$ and $\qcs{L}'$ descend to bounded quasicharacter sheaves on $G$: $B_G(\qcs{L}_0,\mu_0) \iso \qcs{L}$ and $B_G(\qcs{L}'_0,\mu'_0) \iso \qcs{L}'$.
Since bounded quasicharacter sheaves enjoy property \cite{laumon:87a}*{1.1.3.3}, \todo{Justify} it follows (arguing as in \cite{laumon:87a}*{1.1.2.1}) that there is an isomorphism $\qcs{L}_0 \to \qcs{L}'_0$. This morphism is compatible with $\mu_0$ and $\mu'_0$, \todo{Justify} hence a morphism of bounded quasicharacter sheaves.
Its image under $B_G$ is the isomorphism $\qcs{L} \to \qcs{L}'$ of quasicharacter sheaves we seek.
\end{proof}

In the introduction to this paper we noted that, when working with connected commutative algebraic groups over finite fields, Deligne's dictionary can be described using Weil sheaves $\qcs{L}$ for which there exists an isomorphism $m^* \qcs{L} \iso \qcs{L} \boxtimes\qcs{L}$, without keeping track of the isomorphism.
\todo{rephrase objective of section}
In this section use that perspective to find a perfect dictionary between flabby quasicharacter sheaves on $G$ and characters of $G(\Fq)$.

>>>>

\section{Main results}\label{sec:main}

In this section we find $\QCiso{G}$ and $\Hom(\qcs{L},\qcs{L}')$ for every smooth commutative group scheme over $\Fq$; note that we found these in Section~\ref{sec:etale} when $G$ is etale.

@@

%Let $G$ be a smooth commutative group scheme over $\Fq$.
In this section we consider a category of sheaves on $G$ obtained by
replacing the Weil sheaf $(\gqcs{L}, \phi)$ on $\bG$ in the definition of a quasicharacter sheaf with an $\ell$-adic local system on $G$ itself.

@@

We will identify $\bQC(G)$ with its image under $B_G$
and say that a quasicharacter sheaf $\qcs{L} \in \QC(G)$ is \emph{bounded} on $G$
if it is isomorphic to some $B_G(\qcs{L}_0, \mu_0)$.
We will see in Corollary~\ref{cor:bounded-and-finite} that bounded quasicharacter sheaves are exactly quasicharacter sheaves with bounded Trace of Frobenius.\todo{Check this, later.}

\subsection{Quasicharacter sheaves on connected commutative algebraic groups}\label{sec:connected}

In general, $\bQC(G)$ is an essentially
proper subcategory of $\QC(G)$. 
However, if $G$ is connected and of finite type, the categories are equivalent, as we now show.

\begin{lemma}\label{lem:bounded_connected}
If $G$ is a connected commutative algebraic group over $\Fq$ then every quasisharacter sheaf on $G$ is finite and $B_G : \bQC(G) \to \QC(G)$ is an equivalence of categories.
\end{lemma}

@@

\begin{proposition}\label{prop:obj_connected}
 If $G$ is a connected, commutative algebraic group over $\Fq$ then
 \[
  \TrFrob{G} : \QCiso{G} \to G(\Fq)^*
 \]
 is an isomorphism of groups.
 Moreover, for every quasicharacter sheaf $\qcs{L}$ on $G$, $\Aut(\qcs{L}) = 1$.
\end{proposition}

\begin{proof}
In Lemma~\ref{lem:bounded_connected} we saw that every quasicharacter sheaf on $G$ descends to bounded quasicharacter sheaf on $G$. 
From here, Deligne's function--sheaf dictionary for connected commutative algebraic groups over finite fields \citelist{\cite{deligne:SGA4.5}*{Sommes trig.}, \cite{laumon:87a}*{1.1.3}}  gives us all we need, as we briefly recall.
The key observation is that, by descent theory, pull-back by the Lang isogeny $\Lang : G\to G$ defines an equivalence of categories from $\bQC(G)$ to the category of $1$-dimensional $\ell$-adic representations $V$ of $G(\Fq)$ equipped with an isomorphism $V \to V\otimes V$ such that the trace of $\Lang^*\qcs{L}$ is $\trFrob{\qcs{L}}^{-1}$ for every bounded quasicharacter sheaf $\qcs{L}$ on $G$. 
Note that since $V$ is $1$-dimensional, the choice of $V \to V\otimes V$ is exactly the choice of an isomorphism $V\iso \EE$.
Using the equivalence given by $\Lang^*$, it follows that if $\Lang^*\qcs{L} = V$ with character $\psi$ and isomorphism $\mu : V\to V\otimes V$ then $\Aut(\qcs{L})$ is the group of automorphisms of $\rho : V\to V$ for which 
\[
\begin{tikzcd}
\arrow{d}{\mu} V \arrow{r}{\rho} & V\arrow{d}{\mu}\\
V\otimes V \arrow{r}{\rho\otimes \rho} & V\otimes V
\end{tikzcd}
\]
commutes. 
Since the only such isomorphism $\rho$ is $\id_V$, it follows that $\Aut(\qcs{L}) = 1$.
\end{proof}

We have just seen that, on a connected commutative algebraic group $G$ over $\Fq$, the category of quasicharacters $\qcs{L}$ on $G$ is equivalent to the category of characters $\psi$ of $G(\Fq)$. This can be used to understand all quasicharacter sheaves on such $G$ using central extensions of the form
\[
\]
Kamgarpour

We have also just seen moreover, that the equality $m^*\psi = \psi \boxtimes \psi$ determines the isomorphism $\mu : \qcs{L} \to \qcs{L}\boxtimes\qcs{L}$ which is part of the data of the quasicharacter sheaf $\qcs{L}$. Means leads us back to a more common perspective on the function--sheaf dictionary in which one considers $1$-dimensional local systems $\qcs{L}$ on $G$ for which \emph{there exists} an isomorphism $m^*\qcs{L} \iso \qcs{L} \boxtimes\qcs{L}$, or sometimes $1$-dimensional local systems $\gqcs{L}$ on $\bG$ for which \emph{there exists} an isomorphism $\Frob{G}^*\gqcs{L} \iso \gqcs{L}$ and an isomorphism $\bm^*\gqcs{L} \iso \gqcs{L} \boxtimes\gqcs{L}$. 
%In the introduction to this paper we noted that, when working with connected commutative algebraic groups over finite fields, Deligne's dictionary can be described using Weil sheaves $\qcs{L}$ for which there exists an isomorphism $m^* \qcs{L} \iso \qcs{L} \boxtimes\qcs{L}$, without keeping track of the isomorphism.
We have seen that the category of quasicharacter sheaves on commutative smooth group schemes over finite fields specializes to  
that notion when $G$ is connected and of finite type.
However, as we have already seen in Section~\ref{sec:etale}, that notion is not up to the not appropriate if one wishes to extend the dictionary to commutative smooth group schemes over finite fields.
